---
title: "Conclusion"
format:
  html: 
    toc: false
    link-external-icon: true
    link-external-newwindow: true
    link-external-filter: ^(?:http:|https:)\/\/www\.openstatprog\.com
---

```{r include = FALSE}
source(here::here("R", "buttons.R"))
```

By completing this module, you've built a solid understanding of R's tool for representing categorical data:
**factors**.

## Hand-In Assignment

::: practice

## Factors Capstone Assignment

Complete each step below in your own R session, then submit your work (see "What to Hand In" below).

1. **Generate.** Create a character vector called `sizes` containing 15 values randomly sampled (with replacement)
   from `{"small", "medium", "large"}`. Then create a factor called `size_factor` from `sizes`.
1. **Inspect.** Print `levels(size_factor)` and `table(size_factor)`. In a comment, explain what each of these two
   commands tells you.
1. **Modify levels.** Relabel the factor's levels to `{"S", "M", "L"}` (in the order that matches your original
   `{"small", "medium", "large"}` levels).
1. **Conceptual explanation.** In a comment, explain what would happen if you ran `size_factor + 1`, and *why*,
   referring specifically to what a factor actually stores internally.
1. **Use-case reasoning.** In a comment, describe one variable from a dataset you're familiar with (real or
   hypothetical) that *should* be represented as a factor, and one that *should not*, and explain your reasoning for
   each.

### What to Hand In

Save all of your commands, in the order you ran them, in a single plain-text or R script file (e.g.
`yourname_factors.R`), including all four required comments.

A teacher reviewing your submission should be able to see: a correctly constructed factor, correct use of `levels()`
and `table()` with an accurate explanation, correct level relabeling, an accurate conceptual explanation of factor
arithmetic, and sound reasoning about when a factor is (and isn't) the right choice.

:::

## Learning Goals

Now that you've completed this module, you should be able to:

1. Describe the characteristics of an R factor.
1. Explain how factors differ from atomic vectors.
1. Generate new factors in R.
1. Access and modify factor levels.
1. Describe important use-cases for Factors, and explain why factors are the best choice in these situations.

## Additional Reading

For more information on the topics we've covered in this module, the following chapter from Hadley Wickham's *R for
Data Science* is an excellent resource.

- [Factors](https://r4ds.hadley.nz/factors.html) in the [Second Edition](https://r4ds.hadley.nz/index.html).

### Advanced Reading

For a much more technical deep-dive into vectors and vector-like structures (including factors), check out the
following chapter from Hadley Wickham's [Advanced R](https://adv-r.hadley.nz/index.html).

- [Vectors](https://adv-r.hadley.nz/vectors-chap.html)

## Next Steps 

Now that you have a good introduction to the fundamental data structures of the R statistical programming language you're ready to start thinking about how to best organize your data analysis projects. In the next module, we'll explore different aspects of effective R workflows, scripts, working
directories, R projects, file paths, and coding style, to help you create reproducible, organized, and efficient
data analysis projects.

```{r}
#| echo: false
#| results: asis

# next_module_button("https://www.openstatprog.com/workflow/index.html")
next_module_button("in_progress.html")
```
