---
title: "Factors"
page-navigation: false
format:
  html: 
    toc: false
---

```{r, echo = FALSE, include = FALSE}
source(here::here("R", "buttons.R"))
```

::: {.column-margin}

:::: completion-time

**About 60–75 minutes**, including practice exercises and knowledge quiz.

::::

:::

Welcome to the **R Factors** module. In this module, we'll cover a particulalry useful class of data objects:
**factors**. R uses factors to represent categorical data, so factors come up quite frequently in real-world data
analysis (i.e., anytime your analysis involves a discrete grouping variable).

## Learning Goals

After completing this module, you will be able to:

1. Describe the characteristics of an R factor.
1. Explain how factors differ from atomic vectors.
1. Generate new factors in R.
1. Access and modify factor levels.
1. Describe important use-cases for factors, and explain why factors are the best choice in these situations.

Click the button below to get started.

```{r}
#| echo: false
#| results: asis
begin_button("/tutorials/factors/index.html")
```
