---
title: "Factors"
page-navigation: false
format:
  html: 
    toc: false
---

```{r, echo = FALSE, include = FALSE}
source(here::here("R", "buttons.R"))
```

::: {.callout-tip}

## ⏱ Estimated Time to Complete

**About 60–75 minutes**, including practice exercises and the knowledge quiz.

:::

In this module, we'll cover R's tool for representing categorical data: **factors**. It builds directly on the vector, matrix, list, and data frame concepts you've already learned in previous modules.

Content is split into three short tutorials:

1. [Factors](/tutorials/factors/index.qmd)
1. [Creating Factors](/tutorials/factors/creating.qmd)
1. [Special Role of Factors](/tutorials/factors/uniqueness.qmd)
1. [When to Use a Factor](/tutorials/factors/use_cases.qmd)

## Learning Goals

After completing this module, you will be able to:

1. Describe the characteristics of an R factor. 
1. Explain how factors differ from atomic vectors. 
1. Generate new factors in R. 
1. Access and modify factor levels. 
1. Describe important use-cases for factors, and explain why factors are the best choice in these situations.

Click the button below to get started.

```{r}
#| echo: false
#| results: asis
begin_button("/tutorials/factors/index.html")
```
