---
title: "Creating Factors"
---

::: {.callout-note}
This page is hands-on: generating factors and working with their levels.
:::

We create factors using the `factor()` function. This function takes a vector of values (typically numeric or
character) where each unique value represents a group and converts this input vector into a factor.

```{r}
(x <- sample(1:3, 20, TRUE))
(f1 <- factor(x))

(y <- rep(c("foo", "bar"), 10))
(f2 <- factor(y))
```

It may not seem like the `factor()` function did much in the examples above, but that's far from true. Regardless of
what type of input data we use to create a factor, the `factor()` function will always induce three consistent features
in the resulting factor object.

- Factors have their own class: "factor".
- Factors have a "levels" attribute that stores the names of the groups represented by the factor.
- Factors use integer vectors to represent the grouping implied by the input vector.
  - Regardless of the input vector's type, the data underlying the new factor will be an integer vector.

## Class

Factors have their own class: "factor".

```{r}
# Factors have their own class
class(x)
class(f1)
class(y)
class(f2)
```

## Attributes

Factors have a "levels" attribute that stores the names of the groups represented by the factor.

Since `x` and `y` are atomic vectors, they don't have any attributes.

```{r}
attributes(x)
attributes(y)
```

But the factors `f1` and `f2` have *levels* and *class* attributes.

```{r}
attributes(f1)
attributes(f2)
```

## Vector Type

Much like a matrix is just a vector with a *dim* attribute, a factor is just an integer vector with *levels* and *class*
attributes.

  - Regardless of the input vector's type, the data underlying the new factor will be an integer vector.

As you'd expect, `x` and `y` have suitable atomic types (namely, integer and character).

```{r}
typeof(x)
typeof(y)
```

But the factors `f1` and `f2` are both integer vectors, at their core: all factors are.

```{r}
typeof(f1)
typeof(f2)
```

::: {.callout-important}

Unlike vectors and matrices, we cannot freely revert a factor to its input vector by deleting the factor-specific
attributes. In all but the most specialized cases, we'll loose information.

:::

Since `x` was originally an integer vector, this process seems to work for `f1`.

```{r}
attributes(f1) <- NULL
f1
identical(x, f1)
```

But we clearly loose information if we try to naively convert `f2` back to a vector.

```{r}
attributes(f2) <- NULL
f2
identical(y, f2)
```

## Factor Levels

The levels name the categories represented by the factor. By default, the `factor()` function will use the unique values
of the input vector to name the levels of the resulting factor.

```{r}
levels(f1)
levels(f2)
```

We can assign our own level names through the `labels` argument.

```{r}
(z <- sample(10:11, 10, TRUE))

(f3 <- factor(z, labels = c("Group 10", "Group 11")))
levels(f3)

table(z, f3)
```

We can also modify the levels of an existing factor.

```{r}
levels(f3) <- c("g10", "g11")
levels(f3)

table(z, f3)
```

::: practice 

Create a length-20 factor with two levels = {"yes", "no"}.

:::: {.panel-tabset}

## {{< var tab_title.editor >}}

```{webr-r}
```

## {{< var tab_title.solution >}}

As usual, there are myriad acceptable solutions. Here are a couple possibilities.

```{r}
factor(sample(c("yes", "no"), 20, TRUE))
factor(sample(1:2, 20, TRUE), labels = c("yes", "no"))
```

::::

:::

## Quick Check

- What three things does `factor()` always add to its input, regardless of what type the input was?
- What's stored under the hood: the category *names*, or numbers that map to those names?

```{webr-r}
```
