---
title: "Creating Factors"
---

We create factors using the `factor()` function. This function takes a vector of values (typically numeric or
character) where each unique value represents a group and converts this input vector into a factor.

Below, we create a length-10 integer vector, `x`, and convert it to a factor called `f1`.

```{r}
(x <- sample(1:3, 10, TRUE))
(f1 <- factor(x))
```

Here, we create a length-10 character vector, `y`, and convert it to a factor called `f2`.

```{r}
(y <- rep(c("foo", "bar"), 10))
(f2 <- factor(y))
```

It may not seem like the `factor()` function did much in the examples above, but that's far from true. Regardless of
what type of input data we use to create a factor, the `factor()` function will always induce three consistent features
in the resulting factor object.

- Factors have their own class: "factor".
- Factors have a "levels" attribute that stores the names of the groups represented by the factor.
- Factors use integer vectors to represent the grouping implied by the input vector.
  - Regardless of the input vector's type, the data underlying the new factor will be an integer vector.

## Class

Factors have their own class: "factor".

Our example input vectors are both atomic vectors: nothing new here.

```{r}
class(x)
class(y)
```

Both `f1` and `f2`, however, are objects of the "factor" class.

```{r}
class(f1)
class(f2)
```

## Attributes

In addition to their class, factors have a "levels" attribute that stores the names of the groups represented by the
factor.

Since `x` and `y` are atomic vectors, they don't have any attributes.

```{r}
attributes(x)
attributes(y)
```

But the factors `f1` and `f2` have *levels* and *class* attributes.

```{r}
attributes(f1)
attributes(f2)
```

## Factor Levels

A factor's levels name the categories represented by that factor. By default, the `factor()` function will use the
unique values of the input vector to name the levels of the resulting factor.

```{r}
levels(f1)
levels(f2)
```

We can assign our own level names through the `labels` argument.

```{r}
(z <- sample(10:11, 10, TRUE))

(f3 <- factor(z, labels = c("Group 10", "Group 11")))
levels(f3)

table(z, f3)
```

We can also modify the levels of an existing factor.

```{r}
levels(f3) <- c("g10", "g11")
levels(f3)

table(z, f3)
```

### Nuances of Factor Level Mapping

Sometimes, the way that the `factor()` function maps level names to the values of the input vector can be a bit
confusing. Consider the following example.

**Predict** what the following code will print.

- Pay attention to the structure of the input vector and the order of the `labels` vector.

```{webr-r}
factor(
  rep(c("foo", "bar", "baz"), 3),
  labels = c("yes", "yes", "no")
)
```

::: explanation

If your prediction was not correct, I'm guessing you were probably expecting R to directly map the values in the
`labels` vector to the values in the input vector and produce something like the following.

```{r}
#| echo: false

factor(
  rep(c("foo", "bar", "baz"), 3),
  labels = c("yes", "no", "yes")
)
```

This discrepancy comes from an intermediate step that R sneaks in when assigning factor levels. R sorts the unique
values of the input vector *before* mapping the level names provided by the `labels` argument. So, the labels `c("yes",
"yes" "no")` are matched with the *sorted* input values `c("bar", "baz", "foo")`.

You may also have expected the duplicated `"yes"` values in the `labels` vector to cause problems, but R actually uses
these duplicates to do something quite useful: *collapse groups*. The duplicated `"yes"` entries in `labels` correspond
to the input values `"bar"` and `"baz"`, so R happily assigns all cases with input values `"bar"` or `"baz"` to the
`"yes"` group.

We can use this behavior to combine groups when creating new factors or modifying the levels of existing factors.

```{r}
drinks3 <-
	drinks6 <- factor(
		sample(c("coffee", "tea", "water", "juice", "beer", "wine"), 20, TRUE)
	)
levels(drinks3) <-
	c("alcoholic", "caffeinated", "boring", "caffeinated", "boring", "alcoholic")

table(drinks3, drinks6)
```

:::

## Vector Type

Much like a matrix is just a vector with a *dim* attribute, a factor is just an integer vector with *levels* and *class*
attributes.

  - Regardless of the input vector's type, the data underlying the new factor will be an integer vector.

As you'd expect, `x` and `y` have suitable atomic types (namely, integer and character).

```{r}
typeof(x)
typeof(y)
```

But the factors `f1` and `f2` are both integer vectors, at their core: all factors are.

```{r}
typeof(f1)
typeof(f2)
```

::: {.callout-warning}

Unlike vectors and matrices, we cannot freely revert a factor to its input vector by deleting the factor-specific
attributes. In all but the most specialized cases, we'll loose information.

Since `x` was originally an integer vector, this process seems to work for `f1`.

```{r}
attributes(f1) <- NULL
f1
identical(x, f1)
```

But we clearly loose information if we try to naively convert `f2` back to a vector.

```{r}
attributes(f2) <- NULL
f2
identical(y, f2)
```

:::

::: practice 

Create a length-20 factor with two levels = {"yes", "no"}.

:::: {.panel-tabset}

## {{< var tab_title.editor >}}

```{webr-r}
```

## {{< var tab_title.solution >}}

As usual, there are myriad acceptable solutions. Here are a few possibilities.

When we provide a character vector as input, `factor()` will derive the level names from the unique values of the input
vector.

```{r}
factor(rep(c("yes", "no"), 10))
factor(sample(c("yes", "no"), 20, TRUE))
```

Below, we provide various flavors of atomic vector as input and then use the `labels` argument to tell `factor()` how to map the
integer values to factor levels.

```{r}
factor(sample(1:2, 20, TRUE), labels = c("yes", "no"))
factor(rep(c(3.14, 1.96), each = 10), labels = c("no", "yes"))
```

::::

:::

::: knowledge-check

- What attributes differentiate an integer vector from a factor?
- What's stored under the hood: the category *names*, or numbers that map to those names?

```{webr-r}
```
