---
title: "Special Role of Factors"
---

Factors often cause confusion for new (and not-so-new) data analysts. In particular, the `print()` method for factors
shows the level names when printing factors. So, factors with textual level names (e.g., {"male", "female"}) look much
like character vectors when printed, and factors with numeric level names (e.g., {"0", "1"}) look like numeric vectors.

However, factors are not character vectors, numeric vectors, nor any other type of atomic vector. Factors have their own 
class, *factor*, which tells R functions to interpret factors as encoded categorical variables.

Below, we create two potentially confusing factors.

```{r}
f1 <- factor(sample(0:1, 10, TRUE))
f2 <- factor(sample(c("foo", "bar"), 10, TRUE))
```

When we print `f1`, the result looks a lot like an integer vector.

```{r}
f1
```

But, of course, `f1` is not an integer vector.

```{r}
is.integer(f1)
```

Likewise, when we print `f2`, the result looks a lot like a character vector.

```{r}
f2
```

Unsurprisingly, `f2` is not a character vector.

```{r}
is.character(f2)
```

Actually, neither `f1` nor `f2` are any kind of atomic vector.

```{r}
is.vector(f1)
is.vector(f2)
```

## Limitations

Even though factors are built from integer vectors, we can't do mathematical operations on factors. Since
factors represent discrete categories, not continuous quantities, mathematical operations on factors are usually
meaningless.

```{r}
f1 + 1
mean(f1)
```

In the examples above, the requested operations return missing values, `NA`, because we're asking R to do a nonsensical
calculation. So, the result cannot be meaningfully defined.

By design, treating a factor as if it were numeric produces errors and/or null results. This conservative behavior
ensures that categorical data are not accidentally interpreted as numeric values, which would usually produce misleading
results.

## Features

At this point, factors may sound like more trouble than they're worth, but that's certainly not the case. In real-world
data analysis, factors actually save you a lot of time by automatic the process of creating group codes. As we
demonstrated with the conditional boxplot on the last page, R will automatically generate an appropriate set of group
codes when you use a factor in your analysis. Let's look at another very common example.

Below, we're using the `iris` dataset to estimate a linear regression model.

- The outcome variable, `Petal.Length`, is a numeric vector that contains the average lengths of the petals on 100 iris flowers.
- The predictor variable, `Species`, is a three-level factor that records the species of each plant in the sample.

**Predict** how the model estimated below will be parameterized (i.e., what regression coefficients will be estimated).

If you are not familiar with linear regression, feel free to skip directly to the explanation.

```{webr-r}
out <- lm(Petal.Length ~ Species, data = iris)
summary(out)
```

::: explanation

Notice that the estimated regression model contains *two dummy codes*, even though we only specified *one predictor
variable* in the formula `Petal.Length ~ Species`. That's the key point. Since `Species` is a factor, R first creates a
set of dummy codes to represent the species groups, and then uses those dummy codes as the actual predictors in the
regression model.

If we try to estimate the same model using a numeric representation of species, we get entirely incorrect results.

```{r}
# Create an integer representation of the 'Species' variable
iris$IntegerSpecies <- as.integer(iris$Species)

# Estimate the wrong model
out <- lm(Petal.Length ~ IntegerSpecies, data = iris)
summary(out)
```

Since `IntegerSpecies` is an ordinary numeric vector, R doesn't do anything special. So, this model quantifies species
as a continuously varying attribute of the plant, not a qualitative grouping factor.

Of course, we don't absolutely *need* factors, in this case. To estimate an appropriate model, without factors, we could
create the dummy codes ourselves by doing something like the following.

```{r}
# Create dummy codes to represent the species groups
iris$versicolor <- ifelse(iris$IntegerSpecies == 2, 1, 0)
iris$verginica <- ifelse(iris$IntegerSpecies == 3, 1, 0)

# Estimate the correct model
out <- lm(Petal.Length ~ versicolor + verginica, data = iris)
summary(out)
```

Now, we've estimate an appropriate model, but we had to do a lot more work to get there. Clearly, the factor-based
solution is our best option. Factors are a fundamental part of the R data analysis workflow, so R handles them very
intelligently. Whenever you encounter qualitative variables in your analyses, do yourself a favor: use factors!

:::

::: knowledge-check

- If we define `f1` as `f1 <- factor(1:3)`, what would `f1 * 42` return, and why?
- What are some of the practical advantages of using factors in our analyses?

:::
