---
title: "Special Role of Factors"
---

::: {.callout-note}
This page is entirely conceptual: explaining why factors are neither numeric nor
character vectors, even though they can look like either when printed.
:::

Factors often cause confusion for new (and not-so-new) data analysts. In particular, the `print()` method for factors
shows the level names when printing factors. So, factors with textual level names (e.g., {"male", "female"}) look much
like character vectors when printed, and factors with numeric level names (e.g., {"0", "1"}) look like numeric vectors.
However, factors are not character vectors, numeric vectors, nor any other type of atomic vector: they have their own 
class, `factor`, which tells R functions to treat factors as encoded categorical variables.

```{r}
f1 <- factor(sample(0:1, 10, TRUE))
f2 <- factor(sample(c("foo", "bar"), 10, TRUE))

f1 # looks a lot like an integer vector
f2 # looks like a character vector

# Not numeric or character
is.numeric(f1)
is.integer(f1)
is.character(f1)

is.numeric(f2)
is.integer(f2)
is.character(f2)

# Not any type of vector, at all, actually
is.vector(f1)
is.vector(f2)
```

Similarly, even though factors are built from integer vectors, we can't do mathematical operations on factors. Since
factors represent discrete categories, not continuous quantities, mathematical operations on factors are generally
meaningless.

```{r}
f1 + 1
mean(f1)
```

In the examples above, the requested operations return missing values, `NA`, because we're asking R to do a nonsensical
calculation. So, the result cannot be meaningfully defined. By design, treating a factor as if it were numeric produces
errors and/or null results. This conservative behavior ensures that categorical data are not accidentally interpreted as
numeric values, which will usually produce misleading results.

::: {.callout-tip}

## Why This Matters

It's tempting to treat this page as "just more syntax," but the real skill here is conceptual: being able to
*predict* when R will refuse to do something with a factor, and *explain why*, before you ever hit the error
yourself. The Quick Check below and the quiz for this tutorial both test this directly, not just whether you can run
the code.

:::

## Quick Check

- If `f1 <- factor(c("1", "2", "3"))`, what would `f1 + 1` return, and why?
- A colleague says: "I printed my factor and it looked exactly like a numeric vector, so I assume it *is* one." What
  would you tell them?
